\chapter{Christ in the Covenant, the Sabbath, and the Jubilee}

\textit{Theme: Rest, Redemption, and Restoration --- Christ the Lord of the Sabbath and the Proclaimer of Jubilee.}

\section{Introduction: The Rhythm of Divine Rest}

Leviticus 25--26 concludes the covenantal section by revealing three theological realities:

\begin{enumerate}
\item \textbf{Rest (Sabbath and Sabbatical Year)} --- rest from labor and trust in divine provision.
\item \textbf{Redemption (Jubilee)} --- liberty and restoration by grace.
\item \textbf{Relationship (Covenant Blessings)} --- fellowship with God conditioned upon obedience, fulfilled in Christ.
\end{enumerate}

The moral pulse of Leviticus beats in the rhythm of \textbf{``Rest and Release.''} Every seventh day, every seventh year, and every fiftieth year proclaimed that \textbf{God owns the land, the people, and time itself.}

\begin{scripture}
The land is Mine; for you are strangers and sojourners with Me.
\end{scripture}
\scriptureref{Leviticus 25:23}

In Christ, this rhythm reaches its eternal fulfillment:

\begin{itemize}
\item He is \textbf{the true Sabbath} --- rest for the soul (Matt 11:28--29).
\item He is \textbf{the true Redeemer} --- who proclaims the acceptable year of the Lord (Luke 4:18--19).
\item He is \textbf{the true Covenant Mediator} --- securing everlasting blessings (Heb 8:6).
\end{itemize}

\section{Hebrew Word Studies}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lllXX}
\toprule
\textbf{Word} & \textbf{Transliteration} & \textbf{Meaning} & \textbf{Reference} & \textbf{Fulfillment} \\
\midrule
\heb{שַׁבָּת} & \emph{šabbāṯ} & Rest, cessation & Lev 25:2 & Matt 11:28 --- Christ our rest \\
\heb{שַׁבַּת שַׁבָּתוֹן} & \emph{šabbat šabbātôn} & Sabbath of solemn rest & Lev 25:4 & Heb 4:9 --- Eternal rest \\
\heb{יוֹבֵל} & \emph{yōvēl} & Jubilee, ram's horn, liberty & Lev 25:10 & Luke 4:18 --- ``Proclaim liberty'' \\
\heb{גָּאַל} & \emph{gāʾal} & To redeem, buy back & Lev 25:25 & Eph 1:7 --- ``In Him we have redemption'' \\
\heb{בְּרִית} & \emph{berîṯ} & Covenant & Lev 26:9 & Heb 8:6 --- Better covenant \\
\heb{חֶסֶד} & \emph{ḥesed} & Covenant faithfulness / steadfast love & Lev 26:42 & John 1:17 --- Grace and truth \\
\bottomrule
\end{tabularx}
\end{table}

These words together form the grammar of grace: \textbf{rest} (Sabbath), \textbf{release} (Jubilee), and \textbf{relationship} (covenant). All three converge in Christ, who redeems time, land, and people for God.

\section{The Sabbath Year --- Trusting God's Provision (Leviticus 25:1--7)}

Every \textbf{seventh year}, the land rested. The Israelites could not sow, reap, or prune; they lived on what God provided spontaneously.

\begin{scripture}
The Sabbath of the land shall provide food for you.
\end{scripture}
\scriptureref{Leviticus 25:6}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Principle} & \textbf{Meaning} & \textbf{Fulfillment in Christ} \\
\midrule
Land rest & Creation's rhythm restored & Christ, Lord of creation \\
Trust in divine supply & Dependence on God, not toil & ``Your Father knows what you need'' (Matt 6:32) \\
Equality among people & All shared freely & Fellowship of the saints (Acts 2:44--45) \\
\bottomrule
\end{tabularx}
\end{table}

The Sabbath year expressed \textbf{faith through rest} --- the outward picture of inward confidence in God. In Christ, this rest becomes spiritual reality:

\begin{quote}
``We who have believed enter into rest.'' --- Hebrews 4:3.
\end{quote}

\section{The Year of Jubilee --- The Trumpet of Liberty (Leviticus 25:8--55)}

\subsection{The Institution of Jubilee}

Every \textbf{fiftieth year}, after seven sabbatical cycles, the trumpet (\emph{yōvēl}) sounded on the \textbf{Day of Atonement} and proclaimed liberty throughout the land.

\begin{scripture}
You shall hallow the fiftieth year, and proclaim liberty throughout all the land to all its inhabitants.
\end{scripture}
\scriptureref{Leviticus 25:10}

This became Israel's \textbf{year of restoration} --- debts forgiven, slaves freed, property returned.

\subsection{The Threefold Liberation}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Area} & \textbf{Action} & \textbf{Spiritual Fulfillment} \\
\midrule
\textbf{Debt} & All debts cancelled & Forgiveness through Christ's blood (Eph 1:7) \\
\textbf{Slavery} & All slaves released & Deliverance from sin's bondage (Rom 6:18) \\
\textbf{Inheritance} & Land restored & Restoration of spiritual inheritance in Christ (Eph 1:11) \\
\bottomrule
\end{tabularx}
\end{table}

The Jubilee trumpet blew on the Day of Atonement --- liberty rooted in atonement. True freedom flows from forgiveness.

\subsection{The Theology of Jubilee}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Symbol} & \textbf{Meaning} & \textbf{Christological Truth} \\
\midrule
Trumpet blast & Announcement of grace & Gospel proclamation \\
Return to possession & Restoration of inheritance & ``Heirs of God'' (Rom 8:17) \\
Release of servants & Redemption by grace & ``If the Son sets you free...'' (John 8:36) \\
No sowing / reaping & Dependence on provision & Grace, not works \\
Liberty year begins at Atonement & Freedom through blood & ``The acceptable year of the Lord'' (Luke 4:18--19) \\
\bottomrule
\end{tabularx}
\end{table}

The Jubilee functioned as both \textbf{social legislation} and \textbf{spiritual prophecy} --- a foretaste of the \textbf{Kingdom age}, when creation will be delivered from bondage (Rom 8:21).

\section{The Kinsman-Redeemer (Leviticus 25:23--55)}

The laws of redemption introduce one of the most beautiful titles of Christ --- \textbf{the Go'el (Redeemer)}.

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Provision} & \textbf{Description} & \textbf{Fulfillment in Christ} \\
\midrule
Redemption of property (25:25) & A near relative could buy back lost land & Christ, our Brother, restores lost inheritance \\
Redemption of persons (25:47--49) & A kinsman could redeem enslaved family & Christ redeems us from slavery to sin \\
Redemption conditioned on relationship & Only a near kinsman qualified & Christ became man to qualify as Redeemer (Heb 2:14--15) \\
\bottomrule
\end{tabularx}
\end{table}

\begin{scripture}
I know that my Redeemer (Go'el) lives.
\end{scripture}
\scriptureref{Job 19:25}

\begin{scripture}
In Him we have redemption through His blood.
\end{scripture}
\scriptureref{Ephesians 1:7}

The book of Ruth illustrates this principle --- Boaz as the \emph{Go'el} foreshadows Christ, the Redeemer who marries His Bride and restores her inheritance.

\section{Chart --- Sabbath Year vs. Jubilee Year}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Feature} & \textbf{Sabbath Year} & \textbf{Jubilee Year} \\
\midrule
Frequency & Every 7th year & Every 50th year \\
Land & Rests & Returns to original owner \\
Debts & Suspended & Cancelled \\
Servants & Rest & Released \\
Spiritual meaning & Rest in God's provision & Restoration in God's grace \\
Fulfillment in Christ & Spiritual rest (Matt 11:28) & Full redemption (Luke 4:18--19) \\
\bottomrule
\end{tabularx}
\end{table}

Together, these years show \textbf{rest and restoration} --- the essence of redemption. The Sabbath year teaches \textbf{faith}; the Jubilee year proclaims \textbf{freedom}.

\section{The Covenant Blessings and Curses (Leviticus 26)}

After the laws of rest and release, God outlines the \textbf{conditions of covenant relationship} --- obedience bringing blessing, rebellion bringing discipline.

\subsection{Blessings for Obedience (26:1--13)}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Blessing} & \textbf{Description} & \textbf{Fulfillment in Christ} \\
\midrule
Rain in season & Fruitfulness and provision & Spiritual fruit (John 15:5) \\
Peace in the land & Security & Peace with God (Rom 5:1) \\
Victory over enemies & Triumph & ``More than conquerors'' (Rom 8:37) \\
God's presence & Tabernacle among them & ``God with us'' (Matt 1:23) \\
Covenant confirmation & Fellowship & New covenant in His blood (Luke 22:20) \\
\bottomrule
\end{tabularx}
\end{table}

\begin{scripture}
I will walk among you, and be your God, and you shall be My people.
\end{scripture}
\scriptureref{Leviticus 26:12}

This verse anticipates \textbf{Immanuel --- God with us}, the incarnate fulfillment of covenant presence.

\subsection{Curses for Disobedience (26:14--39)}

When Israel disobeyed, the land rested by force during exile (26:34--35). Every unkept Sabbath year was repaid --- fulfilled in the Babylonian captivity (2 Chron 36:21).

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Principle} & \textbf{Leviticus} & \textbf{Fulfillment} \\
\midrule
Discipline for sin & Lev 26:18 & Heb 12:6 --- ``Whom the Lord loves He chastens.'' \\
Exile and desolation & Lev 26:33 & National judgment for unbelief \\
Confession and restoration & Lev 26:40--45 & Israel's future repentance and restoration (Rom 11:26) \\
\bottomrule
\end{tabularx}
\end{table}

Even in judgment, mercy prevailed:

\begin{scripture}
Yet for all that... I will not cast them away... I will remember My covenant.
\end{scripture}
\scriptureref{Leviticus 26:44--45}

This steadfast mercy (\emph{ḥesed}) anticipates the \textbf{new covenant grace} revealed in Christ.

\section{Chart --- The Old and New Covenants}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Aspect} & \textbf{Old Covenant (Leviticus 26)} & \textbf{New Covenant (Hebrews 8--10)} \\
\midrule
Basis & Law & Grace \\
Mediator & Moses & Christ \\
Sacrifice & Animal blood & Christ's blood \\
Scope & National (Israel) & Universal (Church) \\
Duration & Temporary & Eternal \\
Sign & Sabbath & Spirit (Eph 1:13) \\
Promise & Earthly blessings & Spiritual blessings \\
Result & Conditional fellowship & Secure relationship \\
\bottomrule
\end{tabularx}
\end{table}

\begin{quote}
``He is the mediator of a better covenant, established on better promises.'' --- Hebrews 8:6.
\end{quote}

\section{The Fulfillment in Christ: Rest, Release, and Relationship}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lXX}
\toprule
\textbf{Theme} & \textbf{Leviticus Shadow} & \textbf{Fulfillment in Christ} \\
\midrule
\textbf{Rest} & Sabbath year & ``Come unto Me and rest.'' (Matt 11:28) \\
\textbf{Release} & Jubilee & ``Proclaim liberty to the captives.'' (Luke 4:18) \\
\textbf{Redemption} & Go'el (kinsman-redeemer) & ``In Him we have redemption.'' (Eph 1:7) \\
\textbf{Relationship} & Covenant fellowship & ``Abide in Me and I in you.'' (John 15:4) \\
\textbf{Restoration} & Return of land & ``All things reconciled'' (Col 1:20) \\
\bottomrule
\end{tabularx}
\end{table}

The final chapters of Leviticus culminate in Christ --- \textbf{the Rest-Giver, the Redeemer, and the Restorer.}

\section{Doctrinal Synthesis}

\begin{enumerate}
\item \textbf{Sabbath typifies salvation rest.}\\
Rest expresses trust, not inactivity --- faith ceases from self-effort.

\item \textbf{Jubilee prefigures full redemption.}\\
Freedom from sin, debt, and bondage begins now and will be completed at Christ's return.

\item \textbf{The Covenant reveals God's faithfulness.}\\
Even when His people fail, He remembers His covenant of mercy.

\item \textbf{Christ unites all three truths.}\\
He gives rest to the weary, liberty to the captive, and covenant relationship to the redeemed.
\end{enumerate}

\begin{scripture}
The Spirit of the Lord is upon Me... to preach deliverance to the captives, to proclaim the acceptable year of the Lord.
\end{scripture}
\scriptureref{Luke 4:18--19}

\section{Reflection and Discussion}

\begin{enumerate}
\item How do Sabbath and Jubilee together portray the fullness of salvation?

\item Why must atonement precede liberty in the Jubilee sequence?

\item How is Christ both our Redeemer and our Rest?

\item What does God's remembrance of His covenant teach about grace and discipline?

\item How can believers live the Jubilee life today --- spiritually free, yet humbly dependent?
\end{enumerate}

\section{Key Hebrew Vocabulary Recap}

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llXll}
\toprule
\textbf{Word} & \textbf{Transliteration} & \textbf{Meaning} & \textbf{Reference} & \textbf{NT Fulfillment} \\
\midrule
\heb{שַׁבָּת} & \emph{šabbāṯ} & Rest & Lev 25:2 & Heb 4:9--10 \\
\heb{יוֹבֵל} & \emph{yōvēl} & Jubilee / liberty & Lev 25:10 & Luke 4:18 \\
\heb{גָּאַל} & \emph{gāʾal} & Redeem & Lev 25:25 & Eph 1:7 \\
\heb{בְּרִית} & \emph{berîṯ} & Covenant & Lev 26:9 & Heb 8:6 \\
\heb{חֶסֶד} & \emph{ḥesed} & Steadfast love & Lev 26:42 & Rom 8:38--39 \\
\bottomrule
\end{tabularx}
\end{table}

\section{Suggested Reading}

\begin{itemize}
\item Andrew Bonar, \emph{A Commentary on Leviticus} (chs. 25--26)
\item F. B. Meyer, \emph{Christ in the Sabbath and Jubilee}
\item C. H. Mackintosh, \emph{Notes on Leviticus --- Covenant and Jubilee}
\item Alfred Edersheim, \emph{The Temple: Its Ministry and Services} (section on Sabbaths and Jubilees)
\item T. Austin-Sparks, \emph{The Law of Rest and the Power of Life in Christ}
\end{itemize}

\section{Closing Meditation}

\begin{scripture}
This shall be a Jubilee to you; and each of you shall return to his possession.
\end{scripture}
\scriptureref{Leviticus 25:10}

\begin{scripture}
If the Son therefore shall make you free, you shall be free indeed.
\end{scripture}
\scriptureref{John 8:36}

Christ is our eternal Jubilee. He releases us from sin's slavery, restores our lost inheritance, and reconciles us to the Father. The Sabbath calls us to rest; the Jubilee to rejoice; the Covenant to relationship.

In Him, \textbf{time itself finds its fulfillment}, for He is both the Beginning and the End --- the Alpha and the Omega.
